Unsupervised Detection and Tracking of Arbitrary Objects with Dependent Dirichlet Process Mixtures

### 2012-03-27 submitted_ijcv_20120327/neiswanger_wood_unsupervisedDetectionWithDDPM.tex
submitted to IJCV. files in submitted_ijcv_20120327 directory

Research related to general detection and tracking of objects tends to focus on one of either extraction, localization, or tracking. Integrating all three tasks in a system for multiple arbitrary objects and diverse video types is not often a primary focus. A few attempts at accomplishing the three tasks in a cohesive manner have been studied in recent years \cite{brostow2006unsupervised, brox2010object, fragkiadaki2011detection, pece_2002}.
%\willie{An example of general detection and tracking, called blob tracking, typically refers to methods that extract data and localize objects in each frame independently. Tracking involves finding blobs with similar locations and appearances over time. The main blob tracking challenges are the difficulty of segmenting neighboring or occluding objects in a given frame \cite{zhao2004tracking} and preventing multiple tracked blobs from merging into a single object \cite{vermaak_2003}. These difficulties stem from performing detection via segmentation on individual frames.}
This paper furthers this line of work by providing a model that could give rise to a number of algorithms to detect arbitrary objects in videos---particularly in cases where frame-by-frame segmentation is difficult, video quality is low, and extraction is noisy---and maintain the isolation of distinct objects during tracking and through occlusion.

@@

\begin{figure*}
  \centering
  \subfloat[]{\label{fig:ants_img}\includegraphics[width=0.33\textwidth]{../img/usbh_frame_723.pdf}} \hspace{0.1mm}
  \subfloat[]{\label{fig:ants_img2}\includegraphics[width=0.33\textwidth]{../img/usbh_frame_724.pdf}} \hspace{0.1mm}
  \subfloat[]{\label{fig:ants_img_framediff}\includegraphics[width=0.323\textwidth]{../img/usbh_framediff.png}}\\
  \subfloat[]{\label{fig:traffic2_img}\includegraphics[width=0.328\textwidth]{../img/traffic2_frame1.png}} \hspace{0.1mm}
  \subfloat[]{\label{fig:traffic2_img2}\includegraphics[width=0.328\textwidth]{../img/traffic2_frame2.png}} \hspace{0.1mm}
  \subfloat[]{\label{fig:traffic2_img_framediff}\includegraphics[width=0.3273\textwidth]{../img/traffic2_framediff.png}}\\
  \subfloat[]{\label{fig:pets2009_img}\includegraphics[width=0.328\textwidth]{../img/pets2009_frame1.png}} \hspace{0.1mm}
  \subfloat[]{\label{fig:pets2009_img2}\includegraphics[width=0.328\textwidth]{../img/pets2009_frame2.png}} \hspace{0.1mm}
  \subfloat[]{\label{fig:pets2009_img_framediff}\includegraphics[width=0.328\textwidth]{../img/pets2009_framediff.png}}\\
  \subfloat[]{\label{fig:longimg_all}\includegraphics[width=1\textwidth]{../img/longimg_7_detail_7.pdf}}
  \caption{Three pairs of consecutive frames and the results produced by taking the pixel-wise frame difference between each pair (a - i). The final image shows the results of frame differencing over a sequence of images (from the PETS2009/2010 dataset.}
  \label{fig:img_and_framediff}
\end{figure*}

@@

% The basic set of observations extracted from video are the three-dimensional points
% \begin{equation}
%   \centering
%   \bold{x} = ( x_{1}, x_{2}, t ) \in \mathbb{R}^{2} \times \{ 1, \ldots, T \}
% \end{equation}
% where $x_{1}$ and $x_{2}$ are the two spatial locations of a pixel and $t$ is the time index. These data are extracted by recording the positions of pixels lying in regions that contain objects. For our implementation in experiments (Section~\ref{sec:experiments}), we desire a data extraction procedure that is as unsophisticated as possible, to verify that we can perform accurate inference on noisy data extracted using a broadly applicable procedure. Consequentially, we carry out simple frame differencing (introduced in Section~\ref{sec:introduction}) for all experiments, to find pixels at locations where there is motion. This data extraction procedure is simple, computationally inexpensive, and able to be applied to a wide range of static, single-camera videos (note that videos used in all experiments were stationary; moving-camera videos require data extraction methods useful for non-stationary video, such as \cite{chien2002efficient, zhang2007moving}). Examples of pixel location data extracted via frame differencing are shown in Figure~\ref{fig:img_and_framediff}(a-j).

@@

% old discussion on incorporating color features

% The frame differencing and color counting extraction yields a set of data $\bold{X} \subset \mathbb{R}^{2} \times \mathbb{Z}_{+}^{V} \times \{1, \ldots, T \}$, where each element $\bold{x} \in \bold{X}$ can be written
% \begin{equation}
% \bold{x} = ( \bold{x}^{s}, \bold{x}^{c}, t ) = ( x^{s_{1}}, x^{s_{2}}, x^{c_{1}}, \ldots, x^{c_{V}}, t )
% \end{equation}
% where $\bold{x}^{s} \in \mathbb{R}^{2}$ is the two dimensional vector of spatial coordinates,  $\bold{x}^{c} \in \mathbb{Z}_{+}^{V}$ is the $V$ dimensional vector of color counts, and $t \in \{1, \ldots, T \}$ is the time index.
% Other features may also be extracted from each pixel to attempt more sophisticated object representations, including pixel intensity values, feature point (such as corner, shape, or edge) locations or spatial characteristics, and texture representations.

% old (long) data extraction - parts grabbed for new version above
% ----------------------------------------------------------------

% We defined the task of extraction to be the unsupervised determination of which spatiotemporal regions of a video constitute objects, and outlined a number of extraction methods in Section~\ref{sec:priorwork}. The most basic dataset that can be gained from extraction is comprised of the three-dimensional points $\bold{x} = ( x_{1}, x_{2}, t ) \in \mathbb{R}^{2} \times \mathbb{Z}_{+}$, where $x_{1}$ and $x_{2}$ are the two spatial dimensions of a pixel contained in a region found during extraction and $t$ is the frame in which the pixel resides. Once this dataset, which we will refer to as a ``baseline extraction dataset'', has been found, additional features may also be extracted from the video. Examples of additional features include color information, pixel intensity values, feature point (such as corner, shape, or edge) locations or spatial characteristics, and texture representations; the main idea is to choose features which are able to be easily extracted or computed, capture variability in the appearance of objects, and are applicable to a wide variety of object types.

% The baseline extraction dataset has the appearance of worm-like clusters that follow the paths of individual objects as they move (as illustrated in Figure~\ref{fig:longimg_all}). Modeling this data with a dependent Dirichlet process mixture model (Section~\ref{sec:gpuddpmixture}) allows us to perform inference (Section~\ref{sec:inference}), which carries out clustering of the data. By choosing this type of model, we aim to provide a clustering result that isolates each worm-like cluster and infers a sequence of distributions for each, which we use to gain localization and tracking results for each object (Section~\ref{sec:inferencetotrackingresults}).

% For implementation in experiments (Section~\ref{sec:experiments}), we desire an extraction procedure that is as unsophisticated as possible, both to gauge the robustness of this method on potentially noisy extraction data and to ensure that the procedure is applicable to a wide range of videos. Consequently, in all experiments we carry out a method called ``frame-differencing'', which locates motion by recording the positions of pixels that have exhibited differences in intensity or value in succesive frames beyond a given threshold. Frame differencing is unsophisticated, computationally inexpensive, able to be applied to a wide range of static, single-camera videos (note that videos used in all experiments were stationary; moving-camera videos require extraction methods that provide accurate foreground segmentation for non-stationary video). We have found that, when correctly implemented, frame differencing is sufficiently general to extract the desired worm-like clusters from a variety of videos containing multiple moving objects. A few examples of this extraction procedure on pairs of consecutive video frames can be found in Figure~\ref{fig:img_and_framediff}(a-i).

% We furthermore wish to capture the color (or grayscale) value of each pixel and incorporate this color information into our model. For each pixel $x = (i, j, t)$ recorded during frame differencing, we specify a square, $L$ pixels in length, centered on $(i, j)$, that selects a set of pixels surrounding $\bold{x}$ in frame $t$. We also specify a scalar color value that is able to be computed for each pixel; this value could, for example, be some function of the red-green-blue (r-g-b) or hue-saturation-value (h-s-v) characteristic of a pixel. The color value of each of the selected pixels surrounding $\bold{x}$ is recorded. Afterwards, the set of possible color values (ie the range of color values to which a pixel may be assigned) is partitioned into $V$ bins, and the number of pixels with a color value lying in each of the bins yields a $V$ dimensional vector of `color counts'. We will refer to this extraction technique as `color counting'.

% The frame differencing and color counting extraction yields a set of data $\bold{X} \subset \mathbb{R}^{2} \times \mathbb{Z}_{+}^{V} \times \{1, \ldots, T \}$, where each element $\bold{x} \in \bold{X}$ can be written
% \begin{equation}
% \bold{x} = ( \bold{x}^{s}, \bold{x}^{c}, t ) = ( x^{s_{1}}, x^{s_{2}}, x^{c_{1}}, \ldots, x^{c_{V}}, t )
% \end{equation}
% where $\bold{x}^{s} \in \mathbb{R}^{2}$ is the two dimensional vector of spatial coordinates,  $\bold{x}^{c} \in \mathbb{Z}_{+}^{V}$ is the $V$ discrete dimensional vector of color counts, and $t \in \{1, \ldots, T \}$ is the discrete time index.

% In the experiments described in Section~\ref{sec:experiments}, the hue component of the h-s-v value is recorded from all pixels that surround each extracted pixel $\bold{x}$ in the manner described above, choosing $L=5$. Additionally, the set of possible hue values is partitioned into 10 bins, and the number of pixels with a hue value lying in each of the 10 bins is recorded to yield the vector of ``color counts'' $\bold{x}^{c} = x^{c_{1}}, \ldots, x^{c_{10}}$. Hue is chosen to represent object color since it has been demonstrated in previous work as a simple representation of object appearance that allows for distinct objects to be well differentiated \cite{perez_2002, raja_1998, mckenna_1999}.

@@

Dirichlet process mixture (DPM) models (Section \ref{sec:dpmixture}) fall under the heading of Bayesian nonparametric models. These models have been widely used in the past decade to perform nonparametric density estimation and cluster analysis. Data extracted from videos comprises spatiotemporal clusters, each corresponding to a distinct object. Consequently, we are interested in using a class of models known as dependent Dirichlet process mixture (DDPM) models (Section~\ref{sec:ddpmixture}), which are particularly useful for estimating the number of latent classes (clusters) in time dependent data. 
%In this work, estimating the number of clusters in video extraction data is equivalent to estimating the number of distinct objects in a video.

@@

In a GPUDDPM, each observation $\bold{x}_{i,t}$ is associated with an assignment variable $c_{i,t}$ that represents its assignment to a cluster $\theta_{k,t}$. 
%In a DPM, the probability of assignment to a cluster $k$ is a function of the cluster size $m_{k,t}$, which is equal to the number of observations that have been assigned to it. In contrast, 
The sizes of clusters in the GPUDDPM increase when observations are assigned to them, and decrease at later time points when these observations become ``unassigned''. We define a distribution over the size of cluster $k$ at time $t$ ($m_{k,t}$), conditioned on the cluster's previous size ($m_{k,t-1}$), the assignments at time $t$ ($c_{1:N_{t}, t}$), and a deletion parameter ($\rho$),
\begin{equation}
\begin{split}
\label{D_distro}
\text{D} &(m_{k,t} | m_{k,t-1}, \hspace{1mm} c_{1:N_{t}, t}, \hspace{1mm} \rho) := \hspace{2mm} m_{k,t-1} \\
&- \text{Binomial}( m_{k,t} | {m_{k,t-1}, \hspace{1mm} \rho}) + \sum_{i=1}^{N_{t}} \mathbb{I}(c_{i,t} = k)
\end{split}
\end{equation}
$\forall k \in \{1, \ldots, K_{t} \}$, where $K_{t}$ is the number of clusters at time $t$ and $\mathbb{I}(c_{i,t} = k)$ is an indicator function whose value is 1 if $c_{i,t} = k$ and 0 otherwise.

@@

Distributions \eqref{D_distro} and \eqref{C_distro} together comprise what is referred to as the ``Generalized Polya Urn'' \cite{caron_2007}. We can now define the GPUDDPM generatively as
\begin{align}
\label{gpuddpm_def}
\begin{split}
m_{k,t} | m_{k,t-1}, \hspace{1mm} c_{1:N_{t}, t}, \hspace{1mm} \rho  &\sim \text{D}(m_{k,t-1}, \hspace{1mm} c_{1:N_{t}, t}, \hspace{1mm} \rho) \\
\theta_{k, t} | \theta_{k, t-1}   & \sim
\begin{cases}
  P(\theta_{k, t} | \theta_{k, t-1}) \hspace{1mm} \text{if} \hspace{1mm} k \leq K_{t} \\ % \in \{ 1, \ldots, K_{t} \} \\
  \mathbb{G}_{0} \hspace{1mm} \text{if} \hspace{1mm} k = K_{t+1}
\end{cases} \\
c_{i,t} | m_{1:K_{t},t}, \hspace{1mm} \alpha  &\sim  \text{C} (m_{1:K_{t},t}, \hspace{1mm} \alpha) \\
\bold{x}_{i,t} | c_{i,t}, \theta_{1:K_{t}, t} &\sim \text{F}(\theta_{c_{i,t}, t})
\end{split}
\end{align}
$\forall$ times $t \in \{1, \ldots, T\}$ and each cluster $k \in \{ 1, \ldots, K_{t} \}$ at time $t$, where we choose application-specific distributions for F, $\mathbb{G}_{0}$, and $P(\theta_{k, t} | \theta_{k, t-1})$ in Section~\ref{sec:modelspecification}. The graphical model associated with this formulation of the GPUDDPM is shown in Figure~\ref{fig:gpuddpm_gm_1}.
\begin{figure}[h]
        \center{\includegraphics[width=80mm]{../img/gpuddp_gm_1.pdf}}
        \caption{\label{fig:gpuddpm_gm_1} Graphical Model of the Generalized Polya Urn dependent Dirichlet process mixture. The observations at time $t$, $\bold{x}_{1:N_{t},t}$, and their associated assignments $c_{1:N_{t},t}$ are denoted respectively as $\bold{x}_{t}$ and $c_{t}$ (likewise for those at time $t-1$).}
\end{figure}

@@

%The parameters chosen for use in experiments (Section~\ref{sec:experiments}) were accepted based on the fact that they yielded reasonable results. Additionally, a sensitivity analysis was carried out for the concentration and deletion parameters ($\alpha$ and $\rho$), where inference was performed for range of both values (Section~\ref{sec:sensitivityanalysis}).

@@

This section details the MCMC sampler used to perform inference.
% which serves as our first algorithm for unsupervised detection and tracking. It iteratively samples from the conditional distribution of each element in our model given all the rest---i.e. implements Gibbs sampling---to generate approximate samples from the posterior distribution of the model. Additionally, we implement the Metropolis-Hastings (MH) algorithm to sample from a few of these conditional distributions. 
A secondary formulation of the GPUDDPM, which we refer to as the ``Deletion Variable Formulation'' (defined in Section~\ref{sec:deletionvariableformulation}), is used here. This formulation is equivalent to the formulation given in Section~\ref{sec:gpuddpm}, but allows for easier sampling.

@@

We can now define the Deletion Variable Formulation of the GPUDDPM generatively as
\begin{align}
\begin{split}
d_{i,t} | \rho  &\sim \text{Geo}(\rho) + t + 1 \\
\theta_{k, t} | \theta_{k, t-1}  &\sim
\begin{cases}
P(\theta_{k,t} | \theta_{k,t-1}) \hspace{1mm} \text{if} \hspace{1mm} k \leq K_{t} \\ % \in \{ 1, \ldots, K_{t} \} \\
\mathbb{G}_{0} \hspace{1mm} \text{if} \hspace{1mm} k = K_{t} + 1
\end{cases} \\
c_{i,t} | \bold{c}_{1:t-1}, \bold{d}_{1:t-1}, \alpha  &\sim  \text{C}(\bold{c}_{1:t-1}, \bold{d}_{1:t-1}, \alpha) \\
\bold{x}_{i,t} | c_{i,t}, \theta_{c_{i,t},t}  &\sim  \text{F}(\theta_{c_{i,t}, t})
\end{split}
\end{align}
$\forall$ times $t \in \{1, \ldots, T\}$ and clusters $k \in \{ 1, \ldots, K_{t} \} $ at time $t$,
where $\bold{c}_{t} = c_{1:N_{t},t}$, $\bold{d}_{t} = d_{1:N_{t},t}$, and the distributions F, $\mathbb{G}_{0}$, and $P(\theta_{k, t} | \theta_{k, t-1})$ are described in Section~\ref{sec:modelspecification}. This formulation of the GPUDDPM is also used by \cite{gasthaus_thesis} and \cite{caron_2007}. The associated graphical model for this formulation is given in Figure~\ref{fig:gpuddpm_gm_2}.
\begin{figure}[h]
        \center{\includegraphics[width=82mm]{../img/gpuddp_gm_3.pdf}}
        \caption{Graphical model of the Deletion Variable Formulation of the GPUDDPM, showing auxiliary variables.}
        \label{fig:gpuddpm_gm_2}
\end{figure}

@@

% from older version where defined a set S

% We define the set of variables, S, to consist of all observations, assignment variables, deletion variables, cluster parameters (for all active clusters), and auxiliary variables (for all active clusters). We can write
% \begin{equation}
% \text{S} = \{ \boldsymbol{x}_{1:T}, \bold{c}_{1:T}, \bold{d}_{1:T}, \Theta_{1:T}, \bold{z}_{1:T} \}
% \end{equation}
% where

@@

%%%%%
% Need to include ratio of Dirichlet also in the above posterior predictive distro of a single observation, and discuss how to compute any parameters in here that aren't yet specified
%%%%%

@@

The posterior distribution over each of the auxiliary variables $z_{k,t,m}$ can be written as
\begin{equation}
\label{stationary_pdf}
P(z_{k,t,m} | \ldots) \propto  P(z_{k,t,m} | \theta_{k,t-1}) P(\theta_{k,t} | z_{k,t,1:M})
\end{equation}
%fw says to remove:
%
%Samples cannot easily be drawn from this distribution. We therefore use the MH algorithm to sample auxiliary variables from the posterior distribution (for each living cluster). The MH algorithm involves constructing a markov chain with the sought posterior as its stationary distribution, and sampling sequentially from this constructed chain to get an approximate posterior sample. 
%
%
% The MH algorithm is outlined in Algorithm 1.%Algorithm~\ref{alg:MH}. \willie{am i citing wrong?}
% \begin{algorithm}[h!]
% \label{alg:MH}
% \caption{Metropolis Hastings Algorithm}
% \begin{algorithmic}[1]
% \STATE $f(x)$ is the stationary PDF from which $N$ samples are desired \hfill $\triangleright$ eq. \eqref{stationary_pdf}
% \STATE $q( \cdot | x_{i-1})$ is the proposal pdf \hfill $\triangleright$ eq. \eqref{proposal_distro}
% \STATE B is the number of burn in samples
% \STATE $x_{0} \leftarrow$ initialize
% %\STATE B $\leftarrow$ \# of burn in samples
% \FOR{$i = 1 : \text{B} + N$}
% \STATE Sample $x^{*} \sim q( \cdot | x_{i-1})$
% \STATE Sample $\alpha \sim \text{Uniform}(0,1)$
% \STATE $r_{accept} \leftarrow \frac{f(x^{*})q(x_{i-1} | x^{*})} {f(x_{i-1})q(x^{*} | x_{i-1})} $ \hfill $\triangleright$  eq. \eqref{accept_ratio}
% \IF{$\alpha < \text{min} \{ 1, r_{accept} \} $}
% \STATE $x_{i} \leftarrow x^{*}$
% \ELSE
% \STATE $x_{i} \leftarrow x_{i-1}$
% \ENDIF
% \ENDFOR
% \STATE Discard samples $x_{1}, \ldots, x_{\text{B}}$
% \STATE \textbf{Output}: Remaining samples $x_{1}, \ldots, x_{N}$
% \end{algorithmic}
% %\caption{Say thing at end}
% \end{algorithm}
%
We sample a new value for all $M$ auxiliary variables, denoted $z_{k,t,m}^{*}$, using MH, with the proposal distribution
\begin{align}
\begin{split}
\label{proposal_distro}
z_{k,t,m}^{*}  &\sim  P(z_{k,t,m} | \theta_{k,t}) \\
&= \mathcal{N}(z_{k,t,m}^{s} | \boldsymbol{\mu}_{k,t}, \Sigma_{k,t}) \mathcal{M}n(z_{k,t,m}^{c} | \bold{p}_{k,t})
\end{split}
\end{align}
and compute the standard MH acceptance ratio, which in this case simplifies to 
\begin{equation}
\label{accept_ratio}
r_{accept} = \frac{P(\theta_{k,t-1} | z_{k,t,m}^{*})}{P(\theta_{k,t-1} | z_{k,t,m})}
\end{equation}
% and choose $z_{k,t,m}^{*}$ as the next sample in the constructed sequence if $\alpha < \text{min}(1, r_{accept})$, where $\alpha$ is a sample from a $\text{Uniform}(0,1)$ distribution.

@@

The sequence of target distributions for the SMC algorithm may be written as
\begin{equation}
\begin{split}
%\pi_{t}(c_{1:N_{1}, 1}, \ldots, c_{1:N_{t}, t}, & \theta_{1:K_{1}, 1}, \ldots, \theta_{1:K_{t}, t}, m_{1:K_{1}, 1}, \ldots, m_{1:K_{t}, t}) = \\
%\pi_{t-1}(c_{1:N_{1}, 1}, \ldots, c_{1:N_{t-1}, t-1}, & \theta_{1:K_{1}, 1}, \ldots, \theta_{1:K_{t-1}, t-1}, m_{1:K_{1}, 1}, \ldots, m_{1:K_{t-1}, t-1}) \\
\pi_{t}(\bold{c}_{1:t}, \Theta_{1:t}, \bold{m}_{1:t}) & =  \pi_{t-1}(\bold{c}_{1:t-1}, \Theta_{1:t-1}, \bold{m}_{1:t-1})\\
& \times \prod_{i=1}^{N_{t}} P(c_{i,t} | \bold{m}_{t}, \Theta_{t}, \bold{c}_{1:t}, \bold{x}_{1:N_{t}}) \\
& \times \prod_{k=1}^{K_{t}} 
\begin{cases}
\begin{split}
P(\theta_{k,t} | \theta_{k,t-1}) \hspace{2mm} & \text{if} \hspace{2mm} k \leq K_{t-1} \\
\mathbb{G}_{0} \hspace{18mm} & \text{if} \hspace{2mm} k > K_{t-1}
\end{split}
\end{cases} \\
& \times \prod_{k=1}^{K_{t}} \text{D}(m_{k,t} | m_{k,t-1}, c_{1:N_{t-1}, t-1}, \rho)
\end{split}
\end{equation}
where $\bold{c}_{t} = c_{1:N_{t},t}$, $\Theta_{t} = \theta_{1:K_{t},t}$, and $\bold{m}_{t} = m_{1:K_{t},t} $, and D is given by \eqref{D_distro}.
%The distributions used in the SMC algorithm are described in detail in the following sections.

>>>>

Research related to general detection and tracking of objects tends to focus on one of either extraction, localization, or tracking. Integrating all three tasks in a system for multiple arbitrary objects and diverse video types is not often a primary focus. A few attempts at accomplishing the three tasks in a cohesive manner have been studied in recent years \cite{brostow2006unsupervised, brox2010object, fragkiadaki2011detection, pece_2002}.
This paper furthers this line of work by providing a model that could give rise to a number of algorithms to detect arbitrary objects in videos---particularly in cases where frame-by-frame segmentation is difficult, video quality is low, and extraction is noisy---and maintain the isolation of distinct objects during tracking and through occlusion.

@@

\begin{figure*}
  \centering
  \subfloat[]{\label{fig:ants_img}\includegraphics[width=0.33\textwidth]{usbh_frame_723.pdf}} \hspace{0.1mm}
  \subfloat[]{\label{fig:ants_img2}\includegraphics[width=0.33\textwidth]{usbh_frame_724.pdf}} \hspace{0.1mm}
  \subfloat[]{\label{fig:ants_img_framediff}\includegraphics[width=0.323\textwidth]{usbh_framediff.png}}\\
  \subfloat[]{\label{fig:traffic2_img}\includegraphics[width=0.328\textwidth]{traffic2_frame1.png}} \hspace{0.1mm}
  \subfloat[]{\label{fig:traffic2_img2}\includegraphics[width=0.328\textwidth]{traffic2_frame2.png}} \hspace{0.1mm}
  \subfloat[]{\label{fig:traffic2_img_framediff}\includegraphics[width=0.3273\textwidth]{traffic2_framediff.png}}\\
  \subfloat[]{\label{fig:pets2009_img}\includegraphics[width=0.328\textwidth]{pets2009_frame1.png}} \hspace{0.1mm}
  \subfloat[]{\label{fig:pets2009_img2}\includegraphics[width=0.328\textwidth]{pets2009_frame2.png}} \hspace{0.1mm}
  \subfloat[]{\label{fig:pets2009_img_framediff}\includegraphics[width=0.328\textwidth]{pets2009_framediff.png}}\\
  \subfloat[]{\label{fig:longimg_all}\includegraphics[width=1\textwidth]{longimg_7_detail_7.pdf}}
  \caption{Three pairs of consecutive frames and the results produced by taking the pixel-wise frame difference between each pair (a - i). The final image shows the results of frame differencing over a sequence of images (from the PETS2009/2010 dataset.}
  \label{fig:img_and_framediff}
\end{figure*}

@@

@@

@@

Dirichlet process mixture (DPM) models (Section \ref{sec:dpmixture}) fall under the heading of Bayesian nonparametric models. These models have been widely used in the past decade to perform nonparametric density estimation and cluster analysis. Data extracted from videos comprises spatiotemporal clusters, each corresponding to a distinct object. Consequently, we are interested in using a class of models known as dependent Dirichlet process mixture (DDPM) models (Section~\ref{sec:ddpmixture}), which are particularly useful for estimating the number of latent classes (clusters) in time dependent data.

@@

In a GPUDDPM, each observation $\bold{x}_{i,t}$ is associated with an assignment variable $c_{i,t}$ that represents its assignment to a cluster $\theta_{k,t}$.
The sizes of clusters in the GPUDDPM increase when observations are assigned to them, and decrease at later time points when these observations become ``unassigned''. We define a distribution over the size of cluster $k$ at time $t$ ($m_{k,t}$), conditioned on the cluster's previous size ($m_{k,t-1}$), the assignments at time $t$ ($c_{1:N_{t}, t}$), and a deletion parameter ($\rho$),
\begin{equation}
\begin{split}
\label{D_distro}
\text{D} &(m_{k,t} | m_{k,t-1}, \hspace{1mm} c_{1:N_{t}, t}, \hspace{1mm} \rho) := \hspace{2mm} m_{k,t-1} \\
&- \text{Binomial}( m_{k,t} | {m_{k,t-1}, \hspace{1mm} \rho}) + \sum_{i=1}^{N_{t}} \mathbb{I}(c_{i,t} = k)
\end{split}
\end{equation}
$\forall k \in \{1, \ldots, K_{t} \}$, where $K_{t}$ is the number of clusters at time $t$ and $\mathbb{I}(c_{i,t} = k)$ is an indicator function whose value is 1 if $c_{i,t} = k$ and 0 otherwise.

@@

Distributions \eqref{D_distro} and \eqref{C_distro} together comprise what is referred to as the ``Generalized Polya Urn'' \cite{caron_2007}. We can now define the GPUDDPM generatively as
\begin{align}
\label{gpuddpm_def}
\begin{split}
m_{k,t} | m_{k,t-1}, \hspace{1mm} c_{1:N_{t}, t}, \hspace{1mm} \rho  &\sim \text{D}(m_{k,t-1}, \hspace{1mm} c_{1:N_{t}, t}, \hspace{1mm} \rho) \\
\theta_{k, t} | \theta_{k, t-1}   & \sim
\begin{cases}
  P(\theta_{k, t} | \theta_{k, t-1}) \hspace{1mm} \text{if} \hspace{1mm} k \leq K_{t} \\ % \in \{ 1, \ldots, K_{t} \} \\
  \mathbb{G}_{0} \hspace{1mm} \text{if} \hspace{1mm} k = K_{t+1}
\end{cases} \\
c_{i,t} | m_{1:K_{t},t}, \hspace{1mm} \alpha  &\sim  \text{C} (m_{1:K_{t},t}, \hspace{1mm} \alpha) \\
\bold{x}_{i,t} | c_{i,t}, \theta_{1:K_{t}, t} &\sim \text{F}(\theta_{c_{i,t}, t})
\end{split}
\end{align}
$\forall$ times $t \in \{1, \ldots, T\}$ and each cluster $k \in \{ 1, \ldots, K_{t} \}$ at time $t$, where we choose application-specific distributions for F, $\mathbb{G}_{0}$, and $P(\theta_{k, t} | \theta_{k, t-1})$ in Section~\ref{sec:modelspecification}. The graphical model associated with this formulation of the GPUDDPM is shown in Figure~\ref{fig:gpuddpm_gm_1}.
\begin{figure}[h]
        \center{\includegraphics[width=80mm]{gpuddp_gm_1.pdf}}
        \caption{\label{fig:gpuddpm_gm_1} Graphical Model of the Generalized Polya Urn dependent Dirichlet process mixture. The observations at time $t$, $\bold{x}_{1:N_{t},t}$, and their associated assignments $c_{1:N_{t},t}$ are denoted respectively as $\bold{x}_{t}$ and $c_{t}$ (likewise for those at time $t-1$).}
\end{figure}

@@

@@

This section details the MCMC sampler used to perform inference.
A secondary formulation of the GPUDDPM, which we refer to as the ``Deletion Variable Formulation'' (defined in Section~\ref{sec:deletionvariableformulation}), is used here. This formulation is equivalent to the formulation given in Section~\ref{sec:gpuddpm}, but allows for easier sampling.

@@

We can now define the Deletion Variable Formulation of the GPUDDPM generatively as
\begin{align}
\begin{split}
d_{i,t} | \rho  &\sim \text{Geo}(\rho) + t + 1 \\
\theta_{k, t} | \theta_{k, t-1}  &\sim
\begin{cases}
P(\theta_{k,t} | \theta_{k,t-1}) \hspace{1mm} \text{if} \hspace{1mm} k \leq K_{t} \\ % \in \{ 1, \ldots, K_{t} \} \\
\mathbb{G}_{0} \hspace{1mm} \text{if} \hspace{1mm} k = K_{t} + 1
\end{cases} \\
c_{i,t} | \bold{c}_{1:t-1}, \bold{d}_{1:t-1}, \alpha  &\sim  \text{C}(\bold{c}_{1:t-1}, \bold{d}_{1:t-1}, \alpha) \\
\bold{x}_{i,t} | c_{i,t}, \theta_{c_{i,t},t}  &\sim  \text{F}(\theta_{c_{i,t}, t})
\end{split}
\end{align}
$\forall$ times $t \in \{1, \ldots, T\}$ and clusters $k \in \{ 1, \ldots, K_{t} \} $ at time $t$,
where $\bold{c}_{t} = c_{1:N_{t},t}$, $\bold{d}_{t} = d_{1:N_{t},t}$, and the distributions F, $\mathbb{G}_{0}$, and $P(\theta_{k, t} | \theta_{k, t-1})$ are described in Section~\ref{sec:modelspecification}. This formulation of the GPUDDPM is also used by \cite{gasthaus_thesis} and \cite{caron_2007}. The associated graphical model for this formulation is given in Figure~\ref{fig:gpuddpm_gm_2}.
\begin{figure}[h]
        \center{\includegraphics[width=82mm]{gpuddp_gm_3.pdf}}
        \caption{Graphical model of the Deletion Variable Formulation of the GPUDDPM, showing auxiliary variables.}
        \label{fig:gpuddpm_gm_2}
\end{figure}

@@

@@

@@

The posterior distribution over each of the auxiliary variables $z_{k,t,m}$ can be written as
\begin{equation}
\label{stationary_pdf}
P(z_{k,t,m} | \ldots) \propto  P(z_{k,t,m} | \theta_{k,t-1}) P(\theta_{k,t} | z_{k,t,1:M})
\end{equation}
We sample a new value for all $M$ auxiliary variables, denoted $z_{k,t,m}^{*}$, using MH, with the proposal distribution
\begin{align}
\begin{split}
\label{proposal_distro}
z_{k,t,m}^{*}  &\sim  P(z_{k,t,m} | \theta_{k,t}) \\
&= \mathcal{N}(z_{k,t,m}^{s} | \boldsymbol{\mu}_{k,t}, \Sigma_{k,t}) \mathcal{M}n(z_{k,t,m}^{c} | \bold{p}_{k,t})
\end{split}
\end{align}
and compute the standard MH acceptance ratio, which in this case simplifies to 
\begin{equation}
\label{accept_ratio}
r_{accept} = \frac{P(\theta_{k,t-1} | z_{k,t,m}^{*})}{P(\theta_{k,t-1} | z_{k,t,m})}
\end{equation}

@@

The sequence of target distributions for the SMC algorithm may be written as
\begin{equation}
\begin{split}
\pi_{t}(\bold{c}_{1:t}, \Theta_{1:t}, \bold{m}_{1:t}) & =  \pi_{t-1}(\bold{c}_{1:t-1}, \Theta_{1:t-1}, \bold{m}_{1:t-1})\\
& \times \prod_{i=1}^{N_{t}} P(c_{i,t} | \bold{m}_{t}, \Theta_{t}, \bold{c}_{1:t}, \bold{x}_{1:N_{t}}) \\
& \times \prod_{k=1}^{K_{t}} 
\begin{cases}
\begin{split}
P(\theta_{k,t} | \theta_{k,t-1}) \hspace{2mm} & \text{if} \hspace{2mm} k \leq K_{t-1} \\
\mathbb{G}_{0} \hspace{18mm} & \text{if} \hspace{2mm} k > K_{t-1}
\end{split}
\end{cases} \\
& \times \prod_{k=1}^{K_{t}} \text{D}(m_{k,t} | m_{k,t-1}, c_{1:N_{t-1}, t-1}, \rho)
\end{split}
\end{equation}
where $\bold{c}_{t} = c_{1:N_{t},t}$, $\Theta_{t} = \theta_{1:K_{t},t}$, and $\bold{m}_{t} = m_{1:K_{t},t} $, and D is given by \eqref{D_distro}.
